---
title: "Xengsort vs. CellRanger Multiplet Classification"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: yes
  pdf_document:
    toc: no
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

## Overview

This notebook evaluates how well **xengsort**-based read classification (see
`README.md` for the full pipeline: raw FASTQs -> xengsort classify ->
graft/host/ambiguous/neither-specific CellRanger runs) identifies multiplet
droplets - barcodes whose reads come from both a human and a mouse cell -
in PDX (patient-derived xenograft) scRNA-seq data, compared against
CellRanger's own single-reference "barnyard" classification
(`gem_classification.csv`, from a CellRanger run against a combined
human+mouse reference).

It compares barcode sets recovered from the species-specific CellRanger
outputs (`_graft_multi`, `_host_multi`, `_ambiguous_multi`, `_neither_multi`)
against CellRanger's per-barcode `call` column, using Euler/Venn diagrams,
read-depth scatterplots, and basic count-matrix summaries.

## Configuration

```{r config}
# EDIT these two values for your dataset before knitting.
# base_dir  - project root containing a "data" subfolder with the
#             CellRanger multi outputs produced by the xengsort pipeline
# base_type - dataset identifier used to name those outputs, e.g.
#             ${base_type}_multi, ${base_type}_graft_multi,
#             ${base_type}_host_multi, ${base_type}_ambiguous_multi,
#             ${base_type}_neither_multi (see README.md, steps 00-06)
base_dir  <- "."
base_type <- "hgmm_6k"
```

# Xengsort vs. CellRanger multiplet classification

Two independent barcode classifications are compared here:

- **Xengsort-based classification**: derived from the CellRanger outputs of
  the xengsort-classified, separately-aligned graft (human-aligned) and host
  (mouse-aligned) FASTQs. A barcode present only in the human-aligned matrix
  is called `GRCh38`, only in the mouse-aligned matrix `GRCm39`, and a
  barcode present in both is called `Multiplet`.
- **CellRanger GEM classification**: CellRanger's own barnyard call
  (`gem_classification.csv`), from a single CellRanger run against a
  combined human+mouse reference.

The code below loads the graft- and host-aligned filtered barcode matrices,
builds the xengsort-based classification, saves it to
`data/<base_type>_graft_multi/xengsort_classification.csv`, and compares it
against CellRanger's GEM classification with proportional Euler diagrams for
each of the three categories (`GRCh38`, `GRCm39`, `Multiplet`).

```{r}
# Load necessary libraries
library(Seurat)
library(tidyverse)
library(eulerr)
library(gridExtra)

# --- 1. Data Loading ---
path_human <- file.path(base_dir, "data", paste0(base_type, "_graft_multi"), "outs", "per_sample_outs", paste0(base_type, "_graft_multi"), "sample_filtered_feature_bc_matrix")
path_mouse <- file.path(base_dir, "data", paste0(base_type, "_host_multi"), "outs", "per_sample_outs", paste0(base_type, "_host_multi"), "sample_filtered_feature_bc_matrix")

data_human <- Read10X(data.dir = path_human)
data_mouse <- Read10X(data.dir = path_mouse)

bc_human <- colnames(data_human)
bc_mouse <- colnames(data_mouse)

# --- 2. Print Counts ---
cat("Barcodes in Human-aligned:", length(bc_human), "\n")
cat("Barcodes in Mouse-aligned:", length(bc_mouse), "\n")

# --- 3. Xengsort Classification Logic ---
grch38_only <- setdiff(bc_human, bc_mouse)
grcm39_only <- setdiff(bc_mouse, bc_human)
multiplets <- intersect(bc_human, bc_mouse)

xengsort_classification <- data.frame(
  barcode = c(grch38_only, grcm39_only, multiplets),
  call = c(rep("GRCh38", length(grch38_only)),
           rep("GRCm39", length(grcm39_only)),
           rep("Multiplet", length(multiplets))),
  stringsAsFactors = FALSE
)

write.csv(xengsort_classification, file.path(base_dir, "data", paste0(base_type, "_graft_multi"), "xengsort_classification.csv"), row.names = FALSE)

# --- 4. Plotting with eulerr ---

# Function to create a horizontal Euler/Venn plot
plot_euler_horizontal <- function(set_list, title) {
  # eulerr automatically calculates the intersections from a list
  fit <- euler(set_list)
  # quantities = TRUE shows the counts inside the circles
  plot(fit, 
       quantities = TRUE, 
       fills = list(fill = c("#66D2D6", "#FBC740"), alpha = 0.6),
       edges = "black",
       main = title)
}

# Plot A: Alignment Overlap (Human vs Mouse)
# This will render horizontally in your knitted document
p1 <- plot_euler_horizontal(list(Human = bc_human, Mouse = bc_mouse), "Alignment Overlap")
print(p1)

# --- 5. Comparison with GEM Classification ---
gem_path <- file.path(base_dir, "data", paste0(base_type, "_multi"), "outs", "per_sample_outs", paste0(base_type, "_multi"), "analysis", "gem_classification.csv")
gem_classification <- read.csv(gem_path, stringsAsFactors = FALSE)

target_calls <- c("GRCh38", "GRCm39", "Multiplet")
comparison_plots <- list()

for (cat_name in target_calls) {
  x_bc <- xengsort_classification$barcode[xengsort_classification$call == cat_name]
  g_bc <- gem_classification$barcode[gem_classification$call == cat_name]
  
  # Generate the plot for this category
  comparison_plots[[cat_name]] <- plot_euler_horizontal(
    list(Xengsort = x_bc, GEM = g_bc), 
    paste("Category:", cat_name)
  )
}

# To show them horizontally in one row in the knitted document:
# grid.arrange works with eulerr objects because they are grid-based
grid.arrange(grobs = comparison_plots, ncol = 3)
```

# Percent plots

To visualize how the two multiplet-calling methods differ across the full
range of sequencing depth, each barcode's total read count (`GRCh38` +
`GRCm39` reads, from the GEM classification) and percent-mouse read
fraction are plotted, with the multiplet barcodes called by each method
(xengsort-based and CellRanger GEM-based) highlighted in red on otherwise
identical background points. This shows whether the two methods flag
multiplets in similar regions of the read-depth / species-mixture space, or
whether one method is systematically more permissive or conservative.

```{r fig.height=3, fig.width=12}
library(ggplot2)
library(patchwork)
library(dplyr)

# --- 1. Data Preparation ---

# Add totalReads and percentMouse to gem_classification
gem_classification <- gem_classification %>%
  mutate(
    totalReads = GRCh38 + GRCm39,
    percentMouse = (GRCm39 / totalReads) * 100
  )

# --- 2. Scatterplot Function ---

plot_multiplet_highlight <- function(df, highlight_barcodes, title) {
  # Create a temporary column for coloring
  df_plot <- df %>%
    mutate(status = ifelse(barcode %in% highlight_barcodes, "Highlight", "Background")) %>%
    # Sort so highlighted points are plotted last (on top)
    arrange(status) 
  
  ggplot(df_plot, aes(x = totalReads, y = percentMouse, color = status)) +
    geom_point(alpha = 0.5, size = 1) +
    scale_color_manual(values = c("Background" = "grey80", "Highlight" = "red")) +
    # scale_x_log10() + # Log scale is better for UMI/Read counts
    theme_minimal() +
    labs(
      title = title,
      x = "Total Reads per Barcode (Log10)",
      y = "% GRCm39 (Mouse) Reads",
      color = "Classification"
    ) +
    theme(legend.position = "none")
}

# --- 3. Generate the Plots ---

# Get multiplet barcodes from each method
xengsort_multiplets <- xengsort_classification$barcode[xengsort_classification$call == "Multiplet"]
gem_multiplets      <- gem_classification$barcode[gem_classification$call == "Multiplet"]

# Create the two plots
p_xeng <- plot_multiplet_highlight(
  gem_classification, 
  xengsort_multiplets, 
  "Multiplets: Xengsort Method"
)

p_gem <- plot_multiplet_highlight(
  gem_classification, 
  gem_multiplets, 
  "Multiplets: Cell Ranger (GEM)"
)

# Display side-by-side
p_xeng + p_gem
```

# Ambiguous and Neither

Xengsort also produces `ambiguous` (reads with support from both genomes)
and `neither` (unclassifiable) read classes, which are separately aligned
to the combined barnyard reference (see steps `03`-`06` of the xengsort
pipeline in `README.md`). This section checks how the barcodes recovered
from those two classes relate to the xengsort-based and CellRanger
GEM-based classifications above, using Euler diagrams of barcode overlap.
Each dataset is processed only if its output folder exists, since not every
run generates a usable `ambiguous` or `neither` class.

```{r}
# Load libraries
library(Seurat)
library(tidyverse)
library(eulerr)
library(gridExtra)

# Construct paths
path_ambiguous <- file.path(base_dir, "data", paste0(base_type, "_ambiguous_multi"), "outs", "per_sample_outs", paste0(base_type, "_ambiguous_multi"), "sample_filtered_feature_bc_matrix")
path_neither   <- file.path(base_dir, "data", paste0(base_type, "_neither_multi"), "outs", "per_sample_outs", paste0(base_type, "_neither_multi"), "sample_filtered_feature_bc_matrix")

# Flags to track existence
exists_ambig   <- dir.exists(path_ambiguous)
exists_neither <- dir.exists(path_neither)

# Helper function for eulerr plots
plot_euler_overlap <- function(set_list, title) {
  fit <- euler(set_list)
  plot(fit, 
       quantities = TRUE, 
       fills = list(fill = c("#A6CEE3", "#B2DF8A"), alpha = 0.6),
       edges = "black",
       main = title)
}

# Initialize plot list and shared barcodes
all_plots <- list()
all_xengsort_bc <- xengsort_classification$barcode
all_gem_bc      <- gem_classification$barcode

# 1. Process Ambiguous if it exists
if (exists_ambig) {
  data_ambiguous <- Read10X(data.dir = path_ambiguous)
  bc_ambiguous <- colnames(data_ambiguous)
  cat("Number of barcodes in Ambiguous dataset:", length(bc_ambiguous), "\n")
  
  all_plots$ambig_xeng <- plot_euler_overlap(list(Ambiguous = bc_ambiguous, Xengsort = all_xengsort_bc), "Ambiguous vs. Xengsort")
  all_plots$ambig_gem  <- plot_euler_overlap(list(Ambiguous = bc_ambiguous, GEM = all_gem_bc), "Ambiguous vs. GEM")
}

# 2. Process Neither if it exists
if (exists_neither) {
  data_neither <- Read10X(data.dir = path_neither)
  bc_neither <- colnames(data_neither)
  cat("Number of barcodes in Neither dataset:", length(bc_neither), "\n")
  
  all_plots$neither_xeng <- plot_euler_overlap(list(Neither = bc_neither, Xengsort = all_xengsort_bc), "Neither vs. Xengsort")
  all_plots$neither_gem  <- plot_euler_overlap(list(Neither = bc_neither, GEM = all_gem_bc), "Neither vs. GEM")
}

# 3. Arrange plots only if any were created
if (length(all_plots) > 0) {
  grid.arrange(grobs = all_plots, 
               ncol = 2, 
               top = "Overlap Analysis: Ambiguous/Neither vs. Previous Classifications")
} else {
  cat("Skipping Overlap Analysis: Neither Ambiguous nor Neither datasets found.\n")
}
```

# Counts

Finally, basic properties of each count matrix (human-aligned,
mouse-aligned, and, if present, ambiguous- and neither-aligned) are
summarized: matrix dimensions, the proportion of zero entries, and the
distribution of non-zero counts.

```{r}
# Function to analyze matrix properties
analyze_matrix_stats <- function(mat, name) {
  cat("\n--- Analysis for:", name, "---\n")
  
  dims <- dim(mat)
  cat("Dimensions (Genes x Barcodes):", dims[1], "x", dims[2], "\n")
  
  total_elements <- as.numeric(dims[1]) * as.numeric(dims[2])
  non_zeros <- length(mat@x)
  prop_zeros <- (total_elements - non_zeros) / total_elements
  cat("Proportion of zeros:", round(prop_zeros * 100, 2), "%\n")
  
  cat("Summary of non-zero count distribution:\n")
  print(summary(mat@x))
  
  invisible(list(dims=dims, sparsity=prop_zeros, summary=summary(mat@x)))
}

# Always analyze Human and Mouse (assuming they exist from previous steps)
analyze_matrix_stats(data_human, "Human-aligned")
analyze_matrix_stats(data_mouse, "Mouse-aligned")

# Conditionally analyze Ambiguous and Neither
if (exists_ambig) {
  analyze_matrix_stats(data_ambiguous, "Ambiguous")
}

if (exists_neither) {
  analyze_matrix_stats(data_neither, "Neither")
}
```
